Experience Report: Haskell in Computational Biology

### 2012-03-06 mrfy_experience_report.tex
a little wordsmithing in the introduction

\documentclass[preprint]{sigplanconf}

@@

@@

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

@@

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

@@

We address this paper to computational biologists who may have an interest in 
functional programming, but this is not an advocacy paper. We do not argue that 
computational biologists \textit{should} use functional programming. Rather, 
argue that computational biologists \textit{can} successfully develop software 
in Haskell, and what to expect if they do.

We also address this paper to the functional programming community, in the 
hopes that our experiences will be enlightening. We believe that our 
experiences can provide lessons about outreach and education.

\subsection{Problem Domain}

Computational biologists write software that answers questions about large data 
sets involving sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). This software comes in two varieties. The first variety 
is \textit{high performance software}, where performance is a top priority; 
this kind of software is usually written in a programming language like C or 
C++. The second variety is software where performance is not a top priority; 
this kind of software written in higher-level languages that emphasize 
convenience and programmer productivity, such as Perl, Python, Ruby, or R.

In this Experience Report, we show that the Haskell programming language is an 
excellent alternative for computational biologists: a Haskell program can 
perform comparably to a C++ program while being as flexible and easy to 
understand as a program written in an interpreted language. Our experience is 
based on an implementation of a novel algorithm for finding remote homologs to 
newly discovered proteins. We call this program MRFy.

@@

\subsection{Our Background}

All three authors are members of the Tufts University Department of Computer 
Science. Noah Daniels has taken a graduate seminar in advanced functional 
programming, which included some Haskell, and spent ten years in industry as a 
professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
has taken a programming languages course heavily focused on functional 
programming, but with no Haskell. Norman Ramsey is the local expert, but he 
wrote no code for this project.

>>>>

\documentclass[preprint,nocopyrightspace,times]{sigplanconf}

@@

\renewcommand\ttdefault{aett}

  \long\def\remark#1{%
      \ifvmode
         \marginpar{\raggedright\hbadness=10000
         \parindent=8pt \parskip=2pt
         \def\baselinestretch{0.8}\tiny
         \itshape\noindent #1\par}%
      \else
          \unskip\raisebox{-3.5pt}{\rlap{$\scriptstyle\diamond$}}%
          \marginpar{\raggedright\hbadness=10000
         \parindent=8pt \parskip=2pt
         \def\baselinestretch{0.8}\tiny
         \itshape\noindent #1\par}%
      \fi}

@@

%\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{Haskell in Computational Biology}   % 'preprint' option specified.

@@

\begin{abstract}
HELLO?  DRAFT OF AN ABSTRACT?
\end{abstract}

% \category{CR-number}{subcategory}{third-level}
% 
% \terms
% term1, term2
% 
% \keywords
% keyword1, keyword2

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins.

\section{The biology}

@@

%%  \subsection{Our Background}
%%  
%%  All three authors are members of the Tufts University Department of Computer 
%%  Science. Noah Daniels has taken a graduate seminar in advanced functional 
%%  programming, which included some Haskell, and spent ten years in industry as a 
%%  professional programmer in languages such as Ruby, C, and C++. Andrew Gallant 
%%  has taken a programming languages course heavily focused on functional 
%%  programming, but with no Haskell. Norman Ramsey is the local expert, but he 
%%  wrote no code for this project.

\remark{Dropped the section on ``our background,'' but the information
  should reappear when and where we discuss barriers to entry in more
  detail. ---NR}
